Plume-SPH 1.0: A three-dimensional, dusty-gas volcanic plume model based on smoothed-particle hydrodynamics

### 2017-03-18 Manuscripts_GMD.tex
modification made based according to dr Pitman's comments

Several newly developed techniques in SPH are adopted and extended to address numerical challenges in simulating multiphase compressible turbulent flow with classical SPH methods. The $SPH-\varepsilon$ turbulence model is adopted to capture mixing at unresolved scales. Heat exchange due to turbulence is then calculated with a Reynolds analogy. A corrected formalism of SPH is used to handle tensile instability and deficiency of particle distribution near the boundaries. We also propose feasible ways to impose velocity inlet and pressure outlet boundary conditions, both of which are scarce in traditional implementations of SPH. The core solver of our model is parallelized with MPI (message passing interface) obtaining good weak scalability and strong scalability. The model is first verified by comparing velocity and concentration distribution along axis and in cross section with experimental results of JPUE (jet or plume that is ejected from a nozzle into a uniform environment). Next, the top height of the Pinatubo eruption of 15 June 1991, as simulated by our model, is consistent with both observations and existing 3D plume models. Profiles of several integrated variables are compared with those calculated in existing 3D plume models, and further verify our model. Analysis of the plume evolution process illustrates that this model is able to reproduce the basic physics of plume development. The comparison also implies that the turbulence model plays a significant role in 3D volcanic plume modeling.
\end{abstract}

\introduction  %% \introduction[modified heading if necessary]
%<<<<<<< HEAD
\subsection{Volcanic ash hazards}
Primary hazards associated with explosive volcanic eruptions include pyroclastic density currents (flows and surges), the widespread deposition of airfall tephra, and the threats to aviation posed by volcanic ash in the atmosphere.
Simulation of all possible hazards with one model is difficult due to diversity among these hazards. Our focus here is the hazards that volcanic ash poses to aircraft and communities.

During volcanic eruptions, VATDs (volcanic ash transport and dispersal models) are used to forecast the location and movement of ash clouds over hours to days.  VATDs use eruption source parameters, such as plume height, mass eruption rate, duration, and the mass fraction distribution of erupted debris finer than about $4 \Phi$ (or $63  \mu$m), which can remain in the cloud for many hours or days. Observational data for such parameters are usually unavailable in the first minutes or hours after an eruption is detected. Moreover, these input parameters are subject to change during an eruption, requiring rapid re-assignment of new parameters. Usually, plume models are used to provide these source terms for VATDs and the forecast accuracy is critically dependent on these models. This paper reports on a new 3D (three dimensional) volcanic plume model designed to exploit the advantages of mesh-free methods for 3D modeling of such plumes that involve multi-phase free boundary flows.

@@

\subsection{Existing plume models}
Several 1D (one dimensional) volcanic plume models have been developed in the past few decades. Ranging from the most basic 1D model \citep{woods1988fluid} which only accounts for mass conservation to more recently developed 1D models  \citep{bursik2001effect, mastin2007user, degruyter2012improving, woodhouse2013interaction, devenish2013using, de2015plume, folch2016fplume, pouget2016sensitivity} which tend to account for more complex physics effects. FPLUME-1.0 \citep{folch2016fplume} accounts for wind effect, entrainment of moisture, water phase change, particle fallout and re-entrainment and even wet aggregation of ash. However, in these 1D models, the entrainment of air is evaluated based on two coefficients: entrainment coefficient for the vertical plume and the entrainment coefficient that describes the effect of wind. Different 1D models adopt different entrainment coefficients based on specific formulation or calibration against well-documented case studies. The feedback from plume to atmosphere is also ignored in 1D models. Even though determination of essential parameters such as the entrainment is not based on first principles, such simple models nevertheless allow us to investigate the importance of physical mechanisms in a volcanic plume. In addition, these simplified models require little computational resource and can run on standard personal computers or on web sites in very short time. As a result, 1D software for volcanic plume development \citep[such as][]{267, 1194, 3541}, combined with VATDs \citep[such as][]{114, draxler2015hysplit} are widely used in research and practice. While these 1D models can generate well-matched results with 3D (three dimensional) models for weak plumes, much greater variability is observed for strong plume scenarios, especially for local variables \citep{costa2016results}. In addition, the need for greater skill in hazards forecasts especially where the plume model is used to generate source conditions for complex VATD models requires 2D (two dimensional) and 3D models.

@@

Recently, a first order, non-equilibrium compressible 3D model, ASHEE \citep{cerminara2016ashee}, was introduced based on three dimensional N phases Eulerian transportation equation. ASHEE is valid for low concentration and low Stokes number region and much faster than N-phase Eulerian model. The model is based an open source numerical solver OpenFOAM adapting its unstructured finite volume solver.

@@

\subsection{Features of SPH}
All of these existing 3D models are mesh based Eulerian methods. To date there are no 3D plume models based on mesh free Lagrangian methods. Lagrangian methods have several features that are suitable for volcanic plume simulation that we outline below. Among such Lagrangian methods, smoothed-particle hydrodynamics \citep{gingold1977smoothed,lucy1977numerical} based simulations have shown good agreement with experiments for many applications in fluid dynamics. And it currently is, by far, the most widely used mesh-free scheme. Our implementations of SPH in volcanology follows on earlier efforts   \citep{bursik2003smoothed,herault2010sph,haddad2016smoothed}.
Specifically, we choose SPH as the numerical method for volcanic plume simulation to enable:
\begin{itemize}
\item Better investigation of mixing phenomena;
\item accurate modeling of the development of ZFE (Zone of Flow Establishment), ZEF (Zone of Established Flow) investigation and relation to column collapse and the questions relating to the development of entrainment; %and proper characterization through numerical experimentation.
\item  easy inclusion of particles of different sizes (phases) and investigation of detailed mechanics of sedimentation and drag force interaction in lower plume.
\end{itemize}
These  will be enabled by the following features of SPH:
\begin{itemize}
\item For mesh based method, either interface tracking (Lagrangian) \citep{harlow1965numerical, wrobel1991computational, cheng1995simplified} or interface capturing (Eulerian) \citep{hirt1981volume, youngs1982time, gerlach2006comparison, gopala2008volume} methods are used to reconstruct the flow interface of free boundary flow. High computational cost, a tendency to form numerical instabilities and the inability to track complex topological changes are the significant drawbacks of tracking techniques \citep{hirt1981volume, unverdi1992front, anderson1998diffuse}. For interface capturing (Eulerian) method, the surrendering of surface detail before the phase transport calculation means that interface reconstruction is required between time steps to recover the interface information, which needs additional numerical effort \citep{hirt1981volume, youngs1982time}. Since SPH is able to adaptively adjust the discretization and automatically construct the interface, SPH does not require additional numerical effort for interface construction and therefore is more suitable for volcanic plume simulation.
\item Advection term in the Navier-Stokes equations does not appear explicitly in discretized formula of SPH (as illustrated in Eq. (\ref{eq:gov-nc-rho}) to Eq. (\ref{eq:gov-nc-e})). So advection term will be treated exactly with SPH \citep{monaghan2005smoothed}.
\item It is easy to include various physics effect (like self gravity, radiative cooling and chemical reaction) into the model. It does not require a major overhaul and re-tooling every time new physics is introduced \citep{monaghan1995sph}. This implies that accounting for more physics is easier for SPH model.
\item With more than one material, each described by its own set of particles, interface problems between phases are often trivial for SPH but difficult for mesh based schemes. So multiphase flow can be easily handled by SPH. Adding of new phases to the model also does not require a major overhaul and re-tooling. As will be shown in later paragraphs, adding of new phases will lead to adding of several lines into the source code for new phases and additional interaction terms between existing phases and newly added phases.
\end{itemize}

@@

Open source availability and  the easy extension features of SPH will facilitate development of a more comprehensive community driven model.

\subsection{Our contributions}
As SPH is a relatively new method in computational fluid dynamics, implementations of SPH for compressible multiphase turbulent flows are few. Several issues endemic to classical SPH necessary for use in this domain are fixed in our implementation. The most popular applications of SPH (and their original motivating application) has been in simulating of free surface flow, such as breaking-waves and flood. Less attention was paid to velocity inlet and pressure outlet boundary conditions which are required in plume modeling.

@@

\section{Physical model} \label{sec:physics-model}
\subsection{Description of the model}
During an explosive eruption, a volcanic jet erupts out from a vent with a speed of several tens to more than 150 meters per second, driven by expanding gas. The jet is initially denser than the surrounding atmosphere and begins to decelerate through negative buoyancy and turbulent interaction with surrounding air. Cauliflower-like vortices are generated along jet margins, within which process, air is entrained and heated up, reducing the bulk density of the entire jet, in many cases, to less than that of the surrounding atmosphere. Once it becomes buoyant, such a jet develops into a plinian or subplinian plume, rising kilometers to tens of kilometers until its heat is exhausted that buoyancy is lost. Jets that lose their momentum before becoming buoyant collapse back onto the ground and transform into pyroclastic flows, surges and ignimbrites. When plume rises up, relatively coarse particle might separate from main stream of the plume falling down onto the ground and probably be re-entrained into the plume at a lower height \citep{ernst1996sedimentation}. Within this process, erupted vapor will condense to liquid (droplet) and even further to ice. Latent heat released from phase change of erupted vapor will further heat up entrained air and make the bulk density further dilute. The entrained vapor might also experience a similar process and impose impact on plume development. Particle aggregation process \citep{carey1982influence,taddeucci2011aggregation}, either due to presence of liquid water, resulting from particle collision or driven by electrostatic forces might occur inside plume and thereby affect the sedimentation.

@@

As an initial effort on exploiting the feasibility and advantage of SPH in plume modeling, our model is designed to describe an injection of well mixed solid and volcanic gas from a circular vent above a flat surface into a stratified stationary atmosphere following SK-3D \citep{suzuki2005numerical}. In this model, molecular viscosity and heat conduction is neglected since turbulent exchange coefficients are dominant. Erupted material consist of solid with different size and mixture of gases (various constituent) is assumed to be well mixed and behave like a single phase fluid (phase 2) which is valid for eruptions with fine particles and ash. Air (also a mixture of different gases) is assumed to be another phase (phase 1). Immediate thermodynamics equilibrium is assumed so that no separate energy equation is needed for each phase. As a result, there is only one energy equation for both phases (heat exchange term between different phases will not show up under this assumption). Immediate dynamics equilibrium is assumed so that no separate momentum equation is needed for each phase. As a result, there is only one vector momentum equation for both phases (drag force term will not show up with this assumption).

Because of the above assumptions, all other microphysical processes (such as the phase changes of \texorpdfstring{H\textsubscript{2}O}, aggregation, disaggregation, absorption of gas on the surface of solids, solution of gas into a liquid) and chemical processes are not considered in this model. These ignored microphysics factors will come to play critical roles under particular eruption conditions. Capturing of these processes needs a more comprehensive model. One critical element in plume development, the effect of wind, is also not considered in this model yet. Introducing of wind effect in our model will need imposing of pressure boundary conditions in a dynamic way, which requires more numerical effort and algorithm design. To summarize, our model is not valid for eruptions where wind effect plays a significant role in its development, usually refered as weak plume. Our model also lacks the ability in modeling plumes with large particles or eruptions in which microphysics plays non-ignorable roles, such as a eruption of \textit{El Chich{\'o}n} volcano on April 4th 1982 \citep{sigurdsson19841982, folch2016fplume}. We are focussed here on developing the SPH based methodoloy in the context of the more basic (and more critical) aspects of volcanic plume and therefore devote our effort to this relative simpler model. It is worthwhile to mention here that because SPH is adopted as our numerical method, adding of these physics into our model would require much less work in terms of programming compared to mesh based methods. Since our plan is an open source distribution of the tool we believe some of these enhancements will rapidly ensue with community participation.

@@

In mesh based methods, governing equations in Eulerian form, Eq. (\ref{eq:gov-cs-rho}) to Eq. (\ref{eq:gov-cs-e}), will be directly discretized. For SPH, governing equations in Lagrange form is needed. By deducting kinetic energy from energy equation, subtracting mass conservation from momentum equation, combining transient term and advection term into material derivative term, the governing equation is put into the final form, in which advection term does not show up explicitly.
\begin{align}
\dfrac{D \rho}{D t} + \rho \nabla \cdot \textbf{v} = 0 \label{eq:gov-nc-rho}\\
\dfrac{D \rho \xi}{D t} + \rho \xi \nabla \cdot \textbf{v} = 0 \label{eq:gov-nc-ks}\\
\dfrac{D \textbf{v}}{D t} + \dfrac{\nabla P}{\rho} =\textbf{g} \label{eq:gov-nc-v}\\
\dfrac{D e}{D t} + \dfrac{P \nabla \cdot \textbf{v}}{\rho} = 0 \label{eq:gov-nc-e}
\end{align}
Governing equations in Lagrangian form, Eq. (\ref{eq:gov-nc-rho}) to Eq.(\ref{eq:gov-nc-e}), and Eulerian form, Eq. (\ref{eq:gov-cs-rho}) to Eq. (\ref{eq:gov-cs-e}), are actually equivalent, as deriving the Lagrangian form from the Eulerian is purely based on mathematics without any additional physical assumptions.

@@

\subsubsection{Non-slip wall boundary}
Velocity is zero for non-slip wall boundary. If we assume the boundary to be adiabatic, heat flux should be zero on the boundary. The flux of mass should also be zero. As a result, internal energy flux, which consists of heat flux and energy flux carried by mass flux, will vanish on the wall boundary. Equation (\ref{eq:wall_bc_rho}) to (\ref{eq:wall_bc_e}) gives no-slip wall boundary condition written in terms of primitive variables.
\begin{align}
\dfrac{\partial \rho}{\partial n} = const = 0\label{eq:wall_bc_rho} \\
\dfrac{\partial \xi}{\partial n} = const = 0 \label{eq:wall_bc_xi}\\ 
\textbf{v} = const =\{0,0,0\}^T \label{eq:wall_bc_v}\\
\dfrac{\partial e }{\partial n} = 0\label{eq:wall_bc_e}
\end{align}

\subsubsection{Open outlet pressure boundary condition}
The pressure of the surrounding atmosphere is given. Except for the pressure, boundary values for density, velocity, and energy on the outlet should depend on the solution. As we ignore the viscosity, the shear stress is ignored and normal stress (whose magnitude equals to pressure) will balance the ambient pressure.
\begin{equation}
p = p_a(h)\label{eq:pressure_bc_p} 
\end{equation}

\section{SPH method} \label{sec:SPH_method}
SPH is a mesh free Lagragian method. In SPH, the domain (control mass) is discretized by a set of particles or discretization points and the position of each particle (discretization point) is updated at every time step based on the motion computed. Approximation of all field variables (velocity, density and pressure, ect.) is obtained by interpolation based on discretization points. The physical laws (such as conservation laws of mass, momentum and energy) written in the form of PDEs (partial differential equations)  or ODEs (ordinary differential equations) need to be transformed into the Lagrangian particle formalism of SPH. Using a kernel function that provides the weighted estimation of the field variables at any point, the integral equations are estimated as sums over particles in a compact subdomain defined by the support of the kernel function defined with respect to the discretization points. Thus, field variables associated to the particle are updated based on its neighbors. Each kernel function has a compact support determined by smoothing length of each particle. There are several review papers by \citet{monaghan1992smoothed, monaghan2005smoothed, rosswog2009astrophysical, price2012smoothed, monaghan2012smoothed}, giving a pretty comprehensive view over SPH, we will only refer here to the representation of the constitutive equations in SPH and put more focus on specific numerical techniques for plume modeling.

>>>>

Several newly developed techniques in SPH are adopted and extended to address numerical challenges in simulating multiphase compressible turbulent flow with classical SPH methods. The $SPH-\varepsilon$ turbulence model is adopted to capture mixing at unresolved scales. Heat exchange due to turbulence is then calculated with a Reynolds analogy. A corrected formalism of SPH is used to handle tensile instability and deficiency of particle distribution near the boundaries. We also propose feasible ways to impose velocity inlet and pressure outlet boundary conditions, both of which are scarce in traditional implementations of SPH. The core solver of our model is parallelized with MPI (message passing interface) obtaining good weak-scalability and strong-scalability. The model is first verified by comparing velocity and concentration distribution along axis and in cross section with experimental results of JPUE (jet or plume that is ejected from a nozzle into a uniform environment). Next, the top height of the Pinatubo eruption of 15 June 1991, as simulated by our model, is consistent with both observations and existing 3D plume models. Profiles of several integrated variables are compared with those calculated in existing 3D plume models, and further verify our model. Analysis of the plume evolution process illustrates that this model is able to reproduce the basic physics of plume development. The comparison also implies that the turbulence model plays a significant role in 3D volcanic plume modeling.
\end{abstract}

\introduction  %% \introduction[modified heading if necessary]
%<<<<<<< HEAD
\subsection{Volcanic ash hazards}
Primary hazards associated with explosive volcanic eruptions include pyroclastic density currents (flows and surges), the widespread deposition of airfall tephra, and the threats to aviation posed by volcanic ash in the atmosphere.
Simulation of all possible hazards with one model is difficult due the fact that different physics of different length scales are dominant in different hazards. Our focus here is the hazards that volcanic ash poses to aircraft.

During volcanic eruptions, VATDs (volcanic ash transport and dispersal models) are used to forecast the location and movement of ash clouds at timescales that range from hours to days. VATDs use eruption source parameters, such as plume height, mass eruption rate, duration, and the mass fraction distribution of erupted debris finer than about $4 \Phi$ (or $63  \mu$m), which can remain in the cloud for many hours or days. Observational data for such parameters are usually unavailable in the first minutes or hours after an eruption is detected. Moreover, these input parameters are subject to change during an eruption, requiring rapid re-assignment of new parameters. Usually, plume models are used to provide these source terms for VATDs and the forecast accuracy is critically dependent on these models. This paper reports on a new 3D (three dimensional) volcanic plume model designed to exploit the advantages of mesh-free methods for 3D modeling of such plumes that involve multi-phase free boundary flows.

@@

\subsection{Existing plume models}
Several 1D (one dimensional) volcanic plume models have been developed in the past few decades, ranging from the most basic 1D model \citep{woods1988fluid} which only accounts for mass conservation to more recently developed 1D models  \citep{bursik2001effect, mastin2007user, degruyter2012improving, woodhouse2013interaction, devenish2013using, de2015plume, folch2016fplume, pouget2016sensitivity} which tend to account for more comprehensive physics effects. FPLUME-1.0 \citep{folch2016fplume} accounts for wind effect, entrainment of moisture, water phase change, particle fallout and re-entrainment and even wet aggregation of ash. However, in these 1D models, the entrainment of air is evaluated based on two coefficients: entrainment coefficient for the vertical plume and the entrainment coefficient that describes the effect of wind. Different 1D models adopt different entrainment coefficients based on specific formulation or calibration against well-documented case studies. The feedback from plume to atmosphere is also ignored in 1D models. Even though determination of essential parameters such as the entrainment is not based on first principles, such simple models nevertheless allow us to investigate the importance of physical mechanisms in a volcanic plume. In addition, these simplified models require little computational resource and can run on standard personal computers or on web sites in very short time. As a result, 1D software for volcanic plume development \citep[such as][]{267, 1194, 3541}, combined with VATDs \citep[such as][]{114, draxler2015hysplit} are widely used in research and practice. While these 1D models can generate well-matched results with 3D (three dimensional) models for weak plumes, much greater variability is observed for strong plume scenarios, especially for local variables \citep{costa2016results}. In addition, the need for greater skill in hazards forecasts especially where the plume model is used to generate source conditions for complex VATD models requires 2D (two dimensional) and 3D models.

@@

Recently, a first order, non-equilibrium compressible 3D model, ASHEE \citep{cerminara2016ashee}, was introduced based on three dimensional N-phases Eulerian transportation equation. ASHEE is valid for low concentration and low Stokes number region and much faster than N-phase Eulerian model. The model is based an open source numerical solver OpenFOAM adapting its unstructured finite volume solver.

@@

\subsection{Features of SPH}
All of these existing 3D models are mesh based Eulerian methods. To date there are no 3D plume models based on mesh free Lagrangian methods. Lagrangian methods have several features that are suitable for volcanic plume simulation that we outline below. Among such Lagrangian methods, smoothed-particle hydrodynamics \citep{gingold1977smoothed,lucy1977numerical} based simulations have shown good agreement with experiments for many applications in fluid dynamics. And it currently is, by far, the most widely used mesh-free scheme. Our implementations of SPH in volcanology follows on earlier efforts   \citep{bursik2003smoothed,herault2010sph,haddad2016smoothed}.
Specifically, we choose SPH as the numerical method for volcanic plume simulation to enable:
\begin{itemize}
\item Better investigation of mixing phenomena;
\item accurate modeling of the development of ZFE (Zone of Flow Establishment), ZEF (Zone of Established Flow) investigation and relation to column collapse and the questions relating to the development of entrainment; %and proper characterization through numerical experimentation.
\item  easy inclusion of particles of different sizes (phases) and investigation of detailed mechanics of sedimentation and drag force interaction in lower plume.
\end{itemize}
These are enabled by the following features of SPH:
\begin{itemize}
\item For mesh based method, either interface tracking (Lagrangian) \citep{harlow1965numerical, wrobel1991computational, cheng1995simplified} or interface capturing (Eulerian) \citep{hirt1981volume, youngs1982time, gerlach2006comparison, gopala2008volume} methods are used to reconstruct the flow interface of free boundary flow. High computational cost, a tendency to form numerical instabilities and the inability to track complex topological changes are the significant drawbacks of tracking techniques \citep{hirt1981volume, unverdi1992front, anderson1998diffuse}. For interface capturing (Eulerian) method, the surrendering of surface detail before the phase transport calculation means that interface reconstruction is required between time steps to recover the interface information, which needs additional numerical effort \citep{hirt1981volume, youngs1982time}. Since SPH is able to adaptively adjust the discretization and automatically construct the interface, SPH does not require additional numerical effort for interface construction and therefore is more suitable for volcanic plume simulation.
\item Advection term in the Navier-Stokes equations does not appear explicitly in discretized formula of SPH (as illustrated in Eq. (\ref{eq:gov-nc-rho}) to Eq. (\ref{eq:gov-nc-e})). So advection term is treated exactly with SPH \citep{monaghan2005smoothed}.
\item It is easy to include various physics effect (like self gravity, radiative cooling and chemical reaction) into the model. It does not require a major overhaul and re-tooling every time new physics is introduced \citep{monaghan1995sph}. This implies that accounting for more physics is easier for SPH model.
\item With more than one material, each described by its own set of particles, interface problems between phases are often trivial for SPH but difficult for mesh based schemes. So multiphase flow can be easily handled by SPH. Adding of new phases to the model also does not require a major overhaul and re-tooling. As will be shown in later paragraphs, adding of new phases only leads to adding of several lines into the source code for new phases and additional interaction terms between existing phases and newly added phases.
\end{itemize}

@@

Open source availability and  the easy extension features of SPH facilitate development of a more comprehensive community driven model.

\subsection{Our contributions}
SPH is still a relatively new method in computational fluid dynamics compared to mesh-based methods, like Finite Volume. Implementations of SPH for compressible multiphase turbulent flows are few. Several issues endemic to classical SPH necessary for use in this domain are fixed in our implementation. The most popular applications of SPH (and their original motivating application) has been in simulating of free surface flow, such as breaking-waves and flood. Less attention was paid to velocity inlet and pressure outlet boundary conditions which are required in plume modeling.

@@

\section{Physical model} \label{sec:physics-model}
\subsection{Description of the model}
During an explosive eruption, a volcanic jet erupts out from a vent with a speed of several tens to more than 150 meters per second, driven by expanding gas. The jet is initially denser than the surrounding atmosphere and begins to decelerate through negative buoyancy and turbulent interaction with surrounding air. Cauliflower-like vortices are generated along jet margins, within which process, air is entrained and heated up, reducing the bulk density of the entire jet, in many cases, to less than that of the surrounding atmosphere. Once it becomes buoyant, such a jet develops into a plinian or subplinian plume, rising up to several kilometers to tens of kilometers until its heat exhausted. Jets that lose their momentum before becoming buoyant collapse back onto the ground and transform into pyroclastic flows, surges and ignimbrites. During the process of plume rising up, relatively larger particles might separate from main stream of the plume, falling down onto the ground and probably be re-entrained into the plume at a lower height \citep{ernst1996sedimentation}. Within this process, erupted vapor condenses to liquid (droplet) and even further to ice. Latent heat released from phase change of erupted vapor  further heats up entrained air and make the bulk density further dilute. The entrained vapor might also experience a similar process and impose impact on plume development. Particle aggregation process \citep{carey1982influence,taddeucci2011aggregation}, either due to presence of liquid water, resulting from particle collision or driven by electrostatic forces might occur inside plume and thereby affect the sedimentation.

@@

As an initial effort on exploiting the feasibility and advantage of SPH in plume modeling, our model is designed to describe an injection of well mixed solid and volcanic gas from a circular vent above a flat surface into a stratified stationary atmosphere following SK-3D \citep{suzuki2005numerical}. In this model, molecular viscosity and heat conduction is neglected since turbulent exchange coefficients are dominant. Erupted material consist of solid with different size and mixture of gases (various constituent) is assumed to be well mixed and behave like a single phase fluid (phase 2) which is valid for eruptions with fine particles and ash. Air (also a mixture of different gases) is assumed to be another phase (phase 1). Immediate thermodynamics equilibrium is assumed so that no separate energy equation is needed for each phase. As a result, there is only one energy equation for both phases (heat exchange term between different phases does not show up under this assumption). Immediate dynamics equilibrium is assumed so that no separate momentum equation is needed for each phase. As a result, there is only one vector momentum equation for both phases (drag force term does not show up with this assumption).

Because of the above assumptions, all other microphysical processes (such as the phase changes of \texorpdfstring{H\textsubscript{2}O}, aggregation, disaggregation, absorption of gas on the surface of solids, solution of gas into a liquid) and chemical processes are not considered in this model. These ignored microphysics factors would play critical roles under particular eruption conditions. Capturing of these processes needs a more comprehensive model. One critical element in plume development, the effect of wind, is also not considered in this model yet. Introducing of wind effect in our model need imposing of pressure boundary conditions in a dynamic way, which requires more numerical effort and algorithm design. To summarize, our model is not valid for eruptions where wind effect plays a significant role in its development, usually refered as weak plume. Our model also lacks the ability in modeling plumes with large particles or eruptions in which microphysics plays non-ignorable roles, such as a eruption of \textit{El Chich{\'o}n} volcano on April 4th 1982 \citep{sigurdsson19841982, folch2016fplume}. We are focussed here on developing the SPH based methodoloy in the context of the more basic (and more critical) aspects of volcanic plume and therefore devote our effort to this relative simpler model. It is worthwhile to mention here that because SPH is adopted as our numerical method, adding of these physics into our model would require much less work in terms of programming compared to mesh based methods. Since our plan is an open source distribution of the tool we believe some of these enhancements will rapidly ensue with community participation.

@@

In mesh based methods, governing equations in Eulerian form, Eq. (\ref{eq:gov-cs-rho}) to Eq. (\ref{eq:gov-cs-e}), are directly discretized. For SPH, governing equations in Lagrange form is needed. By deducting kinetic energy from energy equation, subtracting mass conservation from momentum equation, combining transient term and advection term into material derivative term, the governing equation is put into the final form, in which advection term does not show up explicitly.
\begin{align}
\dfrac{D \rho}{D t} + \rho \nabla \cdot \textbf{v} = 0 \label{eq:gov-nc-rho}\\
\dfrac{D \rho \xi}{D t} + \rho \xi \nabla \cdot \textbf{v} = 0 \label{eq:gov-nc-ks}\\
\dfrac{D \textbf{v}}{D t} + \dfrac{\nabla P}{\rho} =\textbf{g} \label{eq:gov-nc-v}\\
\dfrac{D e}{D t} + \dfrac{P \nabla \cdot \textbf{v}}{\rho} = 0 \label{eq:gov-nc-e}
\end{align}
%Governing equations in Lagrangian form, Eq. (\ref{eq:gov-nc-rho}) to Eq.(\ref{eq:gov-nc-e}), and Eulerian form, Eq. (\ref{eq:gov-cs-rho}) to Eq. (\ref{eq:gov-cs-e}), are actually equivalent, as deriving the Lagrangian form from the Eulerian is purely based on mathematics without any additional physical assumptions.

@@

\subsubsection{Non-slip wall boundary}
Velocity is zero for non-slip wall boundary. If we assume the boundary to be adiabatic, heat flux should be zero on the boundary. The flux of mass should also be zero. As a result, internal energy flux, which consists of heat flux and energy flux carried by mass flux, vanishes on the wall boundary. Equation (\ref{eq:wall_bc_rho}) to (\ref{eq:wall_bc_e}) gives no-slip wall boundary condition written in terms of primitive variables.
\begin{align}
\dfrac{\partial \rho}{\partial n} = const = 0\label{eq:wall_bc_rho} \\
\dfrac{\partial \xi}{\partial n} = const = 0 \label{eq:wall_bc_xi}\\ 
\textbf{v} = const =\{0,0,0\}^T \label{eq:wall_bc_v}\\
\dfrac{\partial e }{\partial n} = 0\label{eq:wall_bc_e}
\end{align}

\subsubsection{Open outlet pressure boundary condition}
The pressure of the surrounding atmosphere is given. Except for the pressure, boundary values for density, velocity, and energy on the outlet should depend on the solution. As we ignore the viscosity, the shear stress is ignored and normal stress (whose magnitude equals to pressure) balances the ambient pressure.
\begin{equation}
p = p_a(h)\label{eq:pressure_bc_p} 
\end{equation}

\section{SPH method} \label{sec:SPH_method}
SPH is a mesh free Lagragian method. In SPH, the domain (control mass) is discretized by a set of particles or discretization points and the position of each particle (discretization point) is updated at every time step based on the motion computed. Approximation of all field variables (velocity, density and pressure, ect.) is obtained by interpolation based on discretization points. The physical laws (such as conservation laws of mass, momentum and energy) written in the form of PDEs (partial differential equations)  or ODEs (ordinary differential equations) need to be transformed into the Lagrangian particle formalism of SPH. Using a kernel function that provides the weighted estimation of the field variables at any point, the integral equations are estimated as sums over particles in a compact subdomain defined by the support of the kernel function defined with respect to the discretization points. Thus, field variables associated to the particle are updated based on its neighbors. Each kernel function has a compact support determined by smoothing length of each particle. There are several review papers by \citet{monaghan1992smoothed, monaghan2005smoothed, rosswog2009astrophysical, price2012smoothed, monaghan2012smoothed}, giving a pretty comprehensive view over SPH, we only refer here to the representation of the constitutive equations in SPH and put more focus on specific numerical techniques for plume modeling.

\subsection{Fundamental principles}
There are several procedures for discretizing governing equations (PDEs or ODEs) with SPH. There is slight difference between discretized formalism obtained by different routes. We present here one of them following \citet{monaghan1992smoothed, monaghan2005smoothed, monaghan2012smoothed}. The starting point of approximating a function with SPH is the translation property of the Dirac function.
\begin{equation}
A(\textbf{x})=\int_{-\infty}^{\infty} A(\textbf{x} \prime) \delta (\textbf{x} \prime - \textbf{x}) d\textbf{x} \prime
\label{eq:Dirac-translation}
\end{equation}
The Dirac function in Eq. (\ref{eq:Dirac-translation}) can be approximated by a weighting function $w(\textbf{x}-\textbf{x}\prime, h)$ (or $w(\textbf{x}\prime-\textbf{x}, h)$) which tends to a Dirac function when $h \rightarrow 0$ :
\begin{equation}
\lim _{h \rightarrow 0} w(\textbf{x} \prime-\textbf{x}, h) =  \delta (\textbf{x} \prime - \textbf{x})
\label{eq:SPH_kernel_delta}
\end{equation}
An arbitrary function $A(\textbf{x})$ can then be approximated by:
\begin{equation}
A(\textbf{x}) \approx <A(\textbf{x})> = \int_{\Omega} A(\textbf{x} \prime) w(\textbf{x}-\textbf{x}\prime, h) d\textbf{x}\prime + O(h^2)
\label{eq:SPH-fundamental-principle}
\end{equation}
Where $h$ is the smoothing length, determining the interaction distance. According to Eq. (\ref{eq:SPH-fundamental-principle}), the order of accuracy of SPH method is second order. However, in practice, second order can not be achieved because there is no guarantee on the symmetry of particle distribution in real simulation \citep{price2012smoothed}.
Recall that $d\textbf{x}\prime = \dfrac{dm \prime}{\rho \prime}$, the integration equation, Eq. (\ref{eq:SPH-fundamental-principle}), can be approximated by summation and lead to an approximation of the function $A$:
\begin{equation}
<A(\textbf{x})> \approx \sum_b m_b \dfrac{A_b}{\rho_b} w(\textbf{x}-\textbf{x}_b, h)
\label{eq:SPH-approximation-sum}
\end{equation}
where the summation is over all the particles within the region of compact support of the weighting function. 
Gradient terms may be straightforwardly calculated by taking the derivative of Eq. (\ref{eq:SPH-approximation-sum}), giving
\begin{equation}
\begin{split}
<\nabla A(\textbf{x})> & = \dfrac{\partial }{\partial \textbf{x}} \int_{\Omega} A(\textbf{x} \prime) w(\textbf{x}-\textbf{x}\prime, h) d\textbf{x}\prime + O(h^2) \\
& \approx \sum_b m_b \dfrac{A_b}{\rho_b} \nabla w(\textbf{x} - \textbf{x}_b, h)
\end{split} 
\label{eq:SPH-scalar-function-gradient}
\end{equation}
For vector quantities the expressions are similar, simply replacing $A$ with $\textbf{A}$ in Eq. (\ref{eq:SPH-approximation-sum}) and Eq. (\ref{eq:SPH-scalar-function-gradient}), giving
\begin{align}
<\textbf{A}(\textbf{x})> \approx \sum_b m_b \dfrac{\textbf{A}_b}{\rho_b} w(\textbf{x}-\textbf{x}_b, h) \\
<\nabla \cdot \textbf{A}(\textbf{x})> \approx \sum_b m_b \dfrac{\textbf{A}_b}{\rho_b} \cdot \nabla w(\textbf{x} - \textbf{x}_b, h) \\
<\nabla \times \textbf{A}(\textbf{x})> \approx \sum_b m_b \dfrac{\textbf{A}_b}{\rho_b} \times \nabla w(\textbf{x} - \textbf{x}_b, h) \\
<\nabla^j \textbf{A}^i(\textbf{x})> \approx \sum_b m_b \dfrac{\textbf{A}_b^i}{\rho_b} \nabla^j w(\textbf{x} - \textbf{x}_b, h) 
\label{eq:SPH-vecctor-function}
\end{align}
%

\subsection{Weighting function}
As described in previous section, the weighting function (or kernel function) is used to replace Dirac function and should satisfy Eq.  (\ref{eq:SPH_kernel_delta}). So it can be viewed as approximation form of Dirac function and satisfy normalization condition:
\begin{equation}
\int	 w(\textbf{x}-\textbf{x}\prime, h) d\textbf{x}\prime = 1
\label{eq:SPH-kernel-normalization-prop}
\end{equation}
Besides normalization, weighting function of particle $a$ has to be symmetric with respect to $a$ to ensure that neighbor particles locate at the same distance away from $a$ contribute equally to SPH summation equation, see Eq. (\ref{eq:SPH-kernel-symmetric}). Weighting function also needs to satisfy conditions such as positivity and compact support. In addition, kernel function must be monotonically decreasing with the distance between particles.\\
\begin{equation}
w(\textbf{x}- \textbf{x} \prime, h) = w(\textbf{x} \prime - \textbf{x}, h)
\label{eq:SPH-kernel-symmetric}
\end{equation}
